---
title: "advanced-eda-plotting"
author: "Michael Spencer"
date: "February 23, 2019"
output: 
  html_document:
    toc: true
---

## Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load libraries and workspace}
# Libraries
library(tigris)
options(tigris_class = "sf")
options(tigris_use_cache = TRUE)
library(sf)
library(tidycensus)
library(tmap)
library(tmaptools) 
library(shinyjs)
library(tidyverse)
library(ggrepel)

census_api_key("891e70ea96814a52b3992d4f85fd4fb500721349")

# Paramaters
  # File Path with County Outcomes
file_path_1 <- "C:\\Users\\Michael Spencer\\Desktop\\ENGR 150 DATA\\c01\\county_outcomes.csv"

  # File Path with County Outcomes Simple
file_path_2 <- "C:\\Users\\Michael Spencer\\Desktop\\ENGR 150 DATA\\c01\\county_outcomes_simple.csv"

  # File Path with County Covariates
file_path_3 <- "C:\\Users\\Michael Spencer\\Desktop\\ENGR 150 DATA\\c01\\cty_covariates.csv"

  # File Path with Commuting Zone Covariates
file_path_4 <- "C:\\Users\\Michael Spencer\\Desktop\\ENGR 150 DATA\\c01\\cz_covariates.csv"

  # File Path with Commuting Zone Outcomes
file_path_5 <- "C:\\Users\\Michael Spencer\\Desktop\\ENGR 150 DATA\\c01\\cz_outcomes.csv"

  # File Path with Simple Commuting Zone Outcomes
file_path_6 <- "C:\\Users\\Michael Spencer\\Desktop\\ENGR 150 DATA\\c01\\cz_outcomes_simple.csv"

  # File Path with National Percentile Outcomes
file_path_7 <- "C:\\Users\\Michael Spencer\\Desktop\\ENGR 150 DATA\\c01\\national_percentile_outcomes.csv"

  # File Path with Percentile to Dollar Crosswalk
file_path_8 <- "C:\\Users\\Michael Spencer\\Desktop\\ENGR 150 DATA\\c01\\pctile_to_dollar_cw.csv"

  # File Path with Census Tract Covariates
file_path_9 <- "C:\\Users\\Michael Spencer\\Desktop\\ENGR 150 DATA\\c01\\tract_covariates.csv"

  # File Path with Census Tract Early Outcomes (parsing errors)
file_path_10 <- "C:\\Users\\Michael Spencer\\Desktop\\ENGR 150 DATA\\c01\\tract_outcomes_early.csv"

  # File Path with Simple Census Tract Outcomes
file_path_11 <- "C:\\Users\\Michael Spencer\\Desktop\\ENGR 150 DATA\\c01\\tract_outcomes_simple.csv"

#Condensed Files
#===============================================================================
county_outcomes_ez_file <- "C:\\Users\\Michael Spencer\\Desktop\\ENGR 150 DATA\\c01\\county_outcomes_ez.csv"

ct_outcomes_ez_file <- "C:\\Users\\Michael Spencer\\Desktop\\ENGR 150 DATA\\c01\\ct_outcomes_ez.csv"

cz_outcomes_ez_file <- "C:\\Users\\Michael Spencer\\Desktop\\ENGR 150 DATA\\c01\\cz_outcomes_ez.csv"

  # Albers projection for 48 contiguous US states
US_ALBERS <- "+proj=aea +lat_1=29.5 +lat_2=45.5 +lat_0=37.5 +lon_0=-96 +x_0=0 +y_0=0 +datum=WGS84 +no_defs"

  # State Fips, Names, & Abbreviations
file_path_state_fips <- "C:\\Users\\Michael Spencer\\Desktop\\ENGR 150 DATA\\c01\\states.csv"

  # File for state borders
file_state_borders <- "C:\\Users\\Michael Spencer\\Desktop\\ENGR 150 DATA\\cb_2017_us_state_20m_sf.rds"

  # File for county borders
file_county_borders <- "C:\\Users\\Michael Spencer\\Desktop\\ENGR 150 DATA\\cb_2017_us_county_20m_sf.rds"

  # Alaska State FIP
alaska_state_fip <- 2

  # Hawaii Stte FIP
hawaii_state_fip <- 15

  # PR State FIP
pr_state_fip <- 72

  # Atlas Colors
atlas_colors <- c("#9a212f", "#9a212f", "#ffffc9", "#31677e", "#31677e")

  # Atlas NA Color
atlas_na_color <- "#d1d1d0"

#===============================================================================

cz_cov <- read_csv(file_path_4)

national_percentile_outcomes <- read_csv(file_path_7)

ct_cov <- read_csv(file_path_9)

county_outcomes_ez <- read_csv(county_outcomes_ez_file)

ct_outcomes_ez <- read_csv(ct_outcomes_ez_file)

cz_outcomes_ez <- read_csv(cz_outcomes_ez_file)

  # Prop White, Non-Educated, etc Color Values
income_colors_values <- 
  c(
    quantile(ct_outcomes_ez$kfr_pooled_pooled_p25, 0, na.rm = TRUE),
    quantile(ct_outcomes_ez$kfr_pooled_pooled_p25, .025, na.rm = TRUE),
    median(ct_outcomes_ez$kfr_pooled_pooled_p25, na.rm = TRUE),
    quantile(ct_outcomes_ez$kfr_pooled_pooled_p25, .975, na.rm = TRUE),
    quantile(ct_outcomes_ez$kfr_pooled_pooled_p25, 1, na.rm = TRUE)
  ) %>% 
  scales::rescale()

  # State fips codes
state_fips <- file_path_state_fips %>% 
  read_csv(col_types = c(col_character(), col_integer(), col_character()))

  # Major cities
major_cities_fips <- tibble(
  city = c("Chicago", "Los Angeles", "San Francisco", "Oakland", "Phoenix", "Louisville"),
  state = c(17, 06, 06, 06, 04, 21),
  state_chr = c("17", "06", "06", "06", "04", "21"),
  county = c(031, 037, 075, 001, 013, 111),
  county_chr = c("031", "037", "075", "001", "013", "111"),
  FIPS = c(17031, 06037, 06075, 06001, 04013, 21111)
)

  # State mapping borders
state_borders <- 
  read_rds(file_state_borders) %>% 
  filter(!(NAME %in% c("Alaska", "Hawaii"))) %>% 
  st_transform(US_ALBERS)

  # County mapping borders
county_borders <-
  read_rds(file_county_borders) %>% 
  filter(!(as.integer(STATEFP) %in% c(alaska_state_fip, hawaii_state_fip))) %>% 
  st_transform(US_ALBERS)

  # Percent formatting function
my_format <- scales::percent_format(accuracy = 1)
```

## Data Quality & Cleaning

```{r data cleaning}
# county_outcomes_ez <-
#   county_outcomes_ez %>% 
#   mutate_at(
#     vars(-state, -county, -hours_wk_pooled_pooled_p25), 
#     ~ ifelse(. < 0, 0, .)
#   ) %>% 
#   mutate_at(
#     vars(-state, -county, -hours_wk_pooled_pooled_p25), 
#     ~ ifelse(. > 1, 1, .)
#   ) %>% 
#   select(-ends_with("_se"))
# 
# ct_outcomes_ez %>% 
#   glimpse()
# 
# ct_outcomes_ez <-
#   ct_outcomes_ez %>% 
#   mutate_at(
#     vars(-state, -county, -tract), 
#     ~ ifelse(. < 0, 0, .)
#   ) %>% 
#   mutate_at(
#     vars(-state, -county, -tract), 
#     ~ ifelse(. > 1, 1, .)
#   ) %>% 
#   select(-ends_with("_se"))
# 
# cz_outcomes_ez <-
#   cz_outcomes_ez %>% 
#   mutate_at(
#     vars(-cz, -czname, -hours_wk_pooled_pooled_p25), 
#     ~ ifelse(. < 0, 0, .)
#   ) %>% 
#   mutate_at(
#     vars(-cz, -czname, -hours_wk_pooled_pooled_p25), 
#     ~ ifelse(. > 1, 1, .)
#   ) %>% 
#   select(-ends_with("_se"))
# 
# write_csv(
#   county_outcomes_ez,
#   "C:\\Users\\Michael Spencer\\Desktop\\ENGR 150 DATA\\c01\\county_outcomes_ez.csv"
# )
# 
# write_csv(
#   ct_outcomes_ez,
#   "C:\\Users\\Michael Spencer\\Desktop\\ENGR 150 DATA\\c01\\ct_outcomes_ez.csv"
# )
# 
# write_csv(
#   cz_outcomes_ez,
#   "C:\\Users\\Michael Spencer\\Desktop\\ENGR 150 DATA\\c01\\cz_outcomes_ez.csv"
# )
```

```{r data renaming}
# ct_names <- names(ct_outcomes_ez)
# 
# county_names <- names(county_outcomes_ez)
# 
# cz_names <- names(cz_outcomes_ez)
# 
# ct_outcomes_ez <-
#   ct_outcomes_ez %>% 
#   rename(
#     teen_birth = "teenbrth_pooled_female_p25",
#     individual_income = "kir_pooled_pooled_p25",
#     male_incarceration_rates = "jail_pooled_pooled_p25",
#     household_income = "kfr_pooled_pooled_p25",
#     married = "married_pooled_pooled_p25",
#     has_dad = "has_dad_pooled_pooled_p25",
#     has_mom = "has_mom_pooled_pooled_p25",
#     has_two_parents = "two_par_pooled_pooled_p25",
#     stay_local = "staytract_pooled_pooled_p25",
#     partners_income = "spouse_rk_pooled_pooled_p25"
#   )
# 
# county_outcomes_ez <- 
#   county_outcomes_ez %>% 
#   rename(
#     teen_birth = "teenbrth_pooled_female_p25",
#     individual_income = "kir_pooled_pooled_p25",
#     male_incarceration_rates = "jail_pooled_pooled_p25",
#     household_income = "kfr_pooled_pooled_p25",
#     married = "married_pooled_pooled_p25",
#     has_dad = "has_dad_pooled_pooled_p25",
#     has_mom = "has_mom_pooled_pooled_p25",
#     has_two_parents = "two_par_pooled_pooled_p25",
#     stay_local = "staytract_pooled_pooled_p25",
#     partner_income = "spouse_rk_pooled_pooled_p25",
#     recieve_assistance_income = "proginc_pooled_pooled_p25",
#     educated_bachelors = "coll_pooled_pooled_p25",
#     educated_hs = "hs_pooled_pooled_p25",
#     educated_graduate_degree = "grad_pooled_pooled_p25",
#     individual_income_native_born_mother = "kir_native_pooled_pooled_p25",
#     individual_income_foreign_born_mother = "kir_imm_pooled_pooled_p25"
#   )
# 
# cz_outcomes_ez <- 
#   cz_outcomes_ez %>% 
#   rename(
#     teen_birth = "teenbrth_pooled_female_p25",
#     individual_income = "kir_pooled_pooled_p25",
#     male_incarceration_rates = "jail_pooled_pooled_p25",
#     household_income = "kfr_pooled_pooled_p25",
#     married = "married_pooled_pooled_p25",
#     has_dad = "has_dad_pooled_pooled_p25",
#     has_mom = "has_mom_pooled_pooled_p25",
#     has_two_parents = "two_par_pooled_pooled_p25",
#     stay_local = "staytract_pooled_pooled_p25",
#     partner_income = "spouse_rk_pooled_pooled_p25",
#     recieve_assistance_income = "proginc_pooled_pooled_p25",
#     educated_bachelors = "coll_pooled_pooled_p25",
#     educated_hs = "hs_pooled_pooled_p25",
#     educated_graduate_degree = "grad_pooled_pooled_p25",
#     individual_income_native_born_mother = "kir_native_pooled_pooled_p25",
#     individual_income_foreign_born_mother = "kir_imm_pooled_pooled_p25"
#   )
```

#===============================================================================

## Summary of Progress

#### Part 1

In this report, I have explored simple outcomes such as adult household income and incarceration rates at the national level by both parental income and race.  Additionally, I have explored how education levels and completion rates vary by parental income and race.  After exploring simple national trends to get a sense of them, I began to dive deeper, looking into trends at the commuting zone level and eventually at the city level.  Though the breadth of trends I looked at wasn't huge, it did give me a sense of some key insights which I'll explore further into progress report 2. In particular, when exploring simple outcomes such as adulthood income rate and incarceration rates within specific cities such as Chicago or Los Angeles, it was striking how much variation there was for census tracts with the same median parental income.  This is a finding I hope to explore more as we go on.

#### Part 2

In this report, I build off of what I have already done by creating static and interactive maps.  To do this, I made several functions which make use of geom_sf and ggplot mapping as well as tmap mapping functions, which are particularly useful for making interactive visuals. In doing so, I've been able to explore how various outcomes, such as income, incaraceration rates, and teen pregnancy varies from census tract to census tract and acros the nation. It would appear that inner city areas are less well off in most of these indicators, which follows with what is already known.

#### Part 3



#===============================================================================

## Functions

### Setting Colors

```{r color finding function}
color_values_ct <- function(small_tracts, outcome, rev_scale) {
  
  vals <-
    c(
      quantile(small_tracts[[outcome]], 0, na.rm = TRUE),
      quantile(small_tracts[[outcome]], .025, na.rm = TRUE) + .00001,
      median(small_tracts[[outcome]], na.rm = TRUE),
      quantile(small_tracts[[outcome]], .975, na.rm = TRUE),
      quantile(small_tracts[[outcome]], 1, na.rm = TRUE)
    ) 
  
  if (rev_scale == 1) vals = rev(vals)
  
  scales::rescale(vals)
  
}
```

```{r tmap color finding function}
color_values_tmap_ct <- function(small_tracts, outcome) {
  
  vals <-
    c(
      quantile(small_tracts[[outcome]], 0, na.rm = TRUE),
      quantile(small_tracts[[outcome]], .025, na.rm = TRUE) + .00001,
      median(small_tracts[[outcome]], na.rm = TRUE),
      quantile(small_tracts[[outcome]], .975, na.rm = TRUE),
      quantile(small_tracts[[outcome]], 1, na.rm = TRUE)
    ) 
  
  vals
  
}
```

```{r tmap color sorter}
color_values <- function(rev_scale) {
  
  if (rev_scale == 1) return(rev(atlas_colors))
  
  atlas_colors
  
}
```

### Creating Labels

```{r legend generator}
legend_generator <- function(outcome) {
  
  case_when(
        str_detect(outcome, "household_income")
        ~ "Adults's mean\nhousehold income\npercentile",
        str_detect(outcome, "incarceration")
        ~ "% of male adults incarcerated\non April 1st, 2010",
        str_detect(outcome, "teen")
        ~ "% of adults who had a\nteen pregnancy",
        TRUE ~ "Outcome Proportion"
      )
  
}
```

### Finding Data

```{r function to find metro data}
metro_bounds <- function(state, metro) {
  
    core_based_statistical_areas(cb = TRUE, year = 2010) %>% 
    filter(grepl(metro, NAME)) %>% 
    st_within(tracts(state = state, year = 2010), .) %>% 
    map_lgl(function(x) {
      if (length(x) == 1) {
        return(TRUE)
      } else {
        return(FALSE)
      }
    }) %>% 
    tracts(state = state, year = 2010)[., ] %>%
    rename("GEOID" = "GEOID10") %>% 
    transform_geo() %>% 
    left_join(
      ct_outcomes_ez, 
      by = c("state", "county", "tract")
    )
  
}
```

```{r function to find state or county data}
# Geography either "county" or "tract"
# For tract level, MUST specify at least the state

county_tract_bounds <- function(geography, state, county) {
  
  bounds <-  
    get_acs(
      geography = geography,
      state = state,
      county = county,
      geometry = TRUE, 
      variables = "B01001_001", 
      year = 2010
    ) %>% 
    transform_geo()
  
  if (geography == "county") {
    return(
      left_join(
        bounds, 
        county_outcomes_ez,
        by = c("state", "county")
      )
    )
  }
  
  if (geography == "tract") {
    return(
      left_join(
        bounds, 
        ct_outcomes_ez,
        by = c("state", "county", "tract")
      )
    )
  }
  
}
```

```{r function to transform projection and mutate fips}
transform_geo <- function(geo_bounds) {
  
  geo_bounds %>% 
    st_transform(US_ALBERS) %>% 
    mutate(
      state = as.integer(str_extract(GEOID, "^\\d{2}")),
      county = as.integer(str_extract(GEOID, "(?<=\\d{2})\\d{3}")),
      tract = as.integer(str_extract(GEOID, "\\d{6}$"))
    ) %>% 
    filter(
      state != alaska_state_fip, 
      state != hawaii_state_fip, 
      state != pr_state_fip
    )
}
```

### Mapping Data

```{r generate tmap}
generate_tmap <- function(bounds, outcome, rev_scale) {
  
  tm_basemap(leaflet::providers$Stamen.TonerLite) +
    tm_shape(bounds) +
    tm_polygons(
      outcome, 
      breaks = color_values_tmap_ct(bounds, outcome),
      labels = my_format(color_values_tmap_ct(bounds, outcome)),
      style = "cont",
      palette = color_values(rev_scale),
      contrast = c(0, 0.9),
      alpha = .8,
      colorNA = atlas_na_color,
      id = "NAME",
      border.alpha = .2,
      title = legend_generator(outcome),
      legend.format = my_format(color_values_tmap_ct(bounds, outcome))
    ) +
    tm_layout(
      frame = FALSE,
      legend.outside = TRUE,
      legend.outside.position = "right",
      legend.format = my_format(color_values_tmap_ct(bounds, outcome)),
      legend.width = Inf,
      legend.title.size = .7
    )
  
}
```

```{r generate geom sf map}
generate_gmap <- function(bounds, outcome, rev_scale) {
  
  ggplot(data = bounds) +
    geom_sf(aes(fill = bounds[[outcome]]), color = NA) +
    coord_sf(datum = NA) +
    theme_void() +
    theme(
      legend.position = "bottom",
      legend.text = element_text(size = 5)
    ) +
    scale_fill_gradientn(
      colors = atlas_colors,
      values = color_values_ct(focused_tracts, outcome, rev_scale),
      na.value = atlas_na_color,
      breaks = seq(.1, 1, by = .1),
      labels =
        c("10%", "20%","30%", "40%","50%", "60%", "70%", "80%", "90%", "100%")
    ) +
    guides(
      fill = 
        guide_colorbar(
          title = legend_generator(outcome),
          title.position = "top",
          title.theme = element_text(size = 7),
          barwidth = 10,
          barheight = .25
        )
    ) 
    # labs(
    #   title = "title",
    #   fill = "hello"
    # )
  
}
```

```{r map metro using tmap}
tmap_metro <- function(state, metro, outcome, rev_scale = 0) {
  
  generate_tmap(metro_bounds(state, metro), outcome, rev_scale)
  
}
```

```{r map region using tmap}
tmap_region <- function(geography, state = NULL, county = NULL, outcome, rev_scale = 0) {
  
  generate_tmap(
    county_tract_bounds(geography, state, county), 
    outcome,
    rev_scale
  )
  
}
```

```{r map region using ggplot}
gmap_region <- function(geography, state = NULL, county = NULL, outcome, rev_scale = 0) {
  
  generate_gmap(
    county_tract_bounds(geography, state, county), 
    outcome, 
    rev_scale
  )
  
}
```

```{r map metro area using ggplot}
gmap_metro <- function(state, metro, outcome, rev_scale = 0) {
  
  generate_gmap(metro_bounds(state, metro), outcome, rev_scale)
  
}
```

```{r map metro area using ggplot v1}
map_metro <- function(state, metro, outcome, rev_scale) {
  
  focused_tracts <-
    core_based_statistical_areas(cb = TRUE, year = 2010) %>% 
    filter(grepl(metro, NAME)) %>% 
    st_within(tracts(state = state, year = 2010), .) %>% 
    map_lgl(function(x) {
      if (length(x) == 1) {
        return(TRUE)
      } else {
        return(FALSE)
      }
    }) %>% 
    tracts(state = state, year = 2010)[., ] %>% 
    st_transform(US_ALBERS) %>% 
    mutate(
      state = as.double(STATEFP10) %% 100,
      county = as.double(COUNTYFP10) %% 10000,
      tract = as.double(TRACTCE10) %% 1000000
    ) %>% 
    left_join(ct_outcomes_ez, by = c("state", "county", "tract"))
  
  ggplot(data = focused_tracts) +
    geom_sf(aes(fill = focused_tracts[[outcome]]), color = NA) +
    coord_sf(datum = NA) +
    theme_void() +
    theme(
      legend.position = "bottom",
      legend.text = element_text(size = 5)
    ) +
    scale_fill_gradientn(
      colors = atlas_colors,
      values = color_values_ct(focused_tracts, outcome, rev_scale),
      na.value = atlas_na_color,
      breaks = seq(.1, 1, by = .1),
      labels =
        c("10%", "20%","30%", "40%","50%", "60%", "70%", "80%", "90%", "100%")
    ) +
    guides(
      fill = 
        guide_colorbar(
          title = legend_generator(outcome),
          title.position = "top",
          title.theme = element_text(size = 7),
          barwidth = 10,
          barheight = .25
        )
    ) +
    labs(
      title = str_c(metro, "Metropolitan Area", sep = " ")
    )
  
}
```

#===============================================================================

## Key Findings and Illustrations

### Plots

#### Exploring National Outcomes

Though the findings below are by no means new, they're neverthless important for understanding our data and its various dimensions.  The plots below demonstrate how incarceration rates and adult household income varies for different races and those who come from higher income backgrounds.  Many of the trends were to be expected - wealthier children grew into wealthier adults and went to jail less frequently.  An aspect to keep in mind while exploring this data further will be how race affects outcomes within specific locations.  As we saw here, race played a huge role in outcomes as white people always had the most preferable outcomes in each scenario, with the occasional exception of Asian-Americans.  Tailing behing them were racial minorities such as Blacks and Hispanics.

```{r adulthood household income rank vs parental household income rank by race}
national_percentile_outcomes %>% 
  ggplot(aes(x = par_pctile)) +
  geom_point(aes(y = kfr_white_pooled, color = "white_ex"), shape = 15, size = 1.5) +
  geom_line(aes(y = s_kfr_white_pooled, color = "white_ex"), size = 1) +
  geom_point(aes(y = kfr_hisp_pooled, color = "hisp_ex"), shape = 16) +
  geom_line(aes(y = s_kfr_hisp_pooled, color = "hisp_ex"), size = 1) +
  geom_point(aes(y = kfr_black_pooled, color = "blk_ex"), shape = 17) +
  geom_line(aes(y = s_kfr_black_pooled, color = "blk_ex"), size = 1) + 
  geom_point(aes(y = kfr_asian_pooled, color = "as_ex"), shape = 17) +
  geom_line(aes(y = s_kfr_asian_pooled, color = "as_ex"), size = 1) +
  scale_y_continuous(
    breaks = seq(.3, .7, by = .1),
    minor_breaks = NULL,
    labels = seq(30, 70, by = 10)
  ) +
  scale_color_manual(
    name = "Race",
    values = c(white_ex = "darkblue", hisp_ex = "orange", blk_ex = "maroon", as_ex = "darkgreen"),
    labels = c(blk_ex = "Black", as_ex = "Asian", white_ex = "White", hisp_ex = "Hispanic")
  ) +
  labs(
    title = "Household Income Rank by Race and Ethnicity",
    subtitle = "Across all races, children growing up in higher income households tend to have\nhigher income households as adults.",
    x = "Parent Household Income Rank",
    y = "Children's Mean Household income Rank (Ages 31-37)"
  ) +
  theme(
    panel.background = element_rect(fill = "white"),
    panel.grid = element_line(color = "gray95")
  ) +
  guides(color = guide_legend(reverse = TRUE))
```

```{r adulthood incarceration rates vs parental household income rank by race}
national_percentile_outcomes %>% 
  ggplot(aes(x = par_pctile)) +
  geom_point(aes(y = jail_white_male, color = "white_ex"), shape = 15, size = 1.5) +
  geom_line(aes(y = s_jail_white_male, color = "white_ex"), size = 1) +
  geom_point(aes(y = jail_hisp_male, color = "hisp_ex"), shape = 16) +
  geom_line(aes(y = s_jail_hisp_male, color = "hisp_ex"), size = 1) +
  geom_point(aes(y = jail_black_male, color = "blk_ex"), shape = 17) +
  geom_line(aes(y = s_jail_black_male, color = "blk_ex"), size = 1) +
  scale_y_continuous(
    breaks = seq(0, .2, by = .05),
    minor_breaks = NULL,
    labels = scales::percent_format(accuracy = 1)
  ) +
  scale_color_manual(
    name = "Race",
    values = c(white_ex = "darkblue", hisp_ex = "orange", blk_ex = "maroon"),
    labels = c("Black", "Hispanic", "White")
  ) +
  labs(
    title = "Male Incarceration Rates by Race and Ethnicity",
    subtitle = "Across all races, children growing up in higher income households tend to have\nlower incarceration rates as adults.",
    x = "Parent Household Income Rank",
    y = "% of Men Incarcerated on April 1, 2010 (Ages 27-32)"
  ) +
  theme(
    panel.background = element_rect(fill = "white"),
    panel.grid = element_line(color = "gray95")
  )
```

```{r HS completion rates vs parental household income}
national_percentile_outcomes %>% 
  ggplot(aes(x = par_pctile)) +
  geom_point(aes(y = hs_white_pooled, color = "white_ex"), shape = 15, size = 1.5) +
  geom_line(aes(y = s_hs_white_pooled, color = "white_ex"), size = 1) +
  geom_point(aes(y = hs_hisp_pooled, color = "hisp_ex"), shape = 16) +
  geom_line(aes(y = s_hs_hisp_pooled, color = "hisp_ex"), size = 1) +
  geom_point(aes(y = hs_black_pooled, color = "blk_ex"), shape = 17) +
  geom_smooth(aes(y = hs_black_pooled, color = "blk_ex"), se = FALSE, size = 1) +
  geom_point(aes(y = hs_asian_pooled, color = "as_ex"), shape = 17) +
  geom_smooth(aes(y = hs_asian_pooled, color = "as_ex"), se = FALSE, size = 1) +
  scale_y_continuous(
    breaks = seq(.5, 1, by = .1),
    minor_breaks = NULL,
    labels = scales::percent_format(accuracy = 1)
  ) +
  scale_color_manual(
    name = "Race",
    values = c(white_ex = "darkblue", hisp_ex = "orange", blk_ex = "maroon", as_ex = "darkgreen"),
    labels = c(blk_ex = "Black", as_ex = "Asian", white_ex = "White", hisp_ex = "Hispanic")
  ) +
  labs(
    title = "National High School Completion Rates by Race and Ethnicity",
    subtitle = "Across all races, children growing up in higher income households tend to have\nhigher high school completion rates.",
    x = "Parent Household Income Rank",
    y = "Children's High School Completion Rates"
  ) +
  theme(
    panel.background = element_rect(fill = "white"),
    panel.grid = element_line(color = "gray95")
  )
```

```{r College completion rates vs parental household income}
national_percentile_outcomes %>% 
  ggplot(aes(x = par_pctile)) +
  geom_point(aes(y = coll_white_pooled, color = "white_ex"), shape = 15, size = 1.5) +
  geom_line(aes(y = s_coll_white_pooled, color = "white_ex"), size = 1) +
  geom_point(aes(y = coll_hisp_pooled, color = "hisp_ex"), shape = 16) +
  geom_line(aes(y = s_coll_hisp_pooled, color = "hisp_ex"), size = 1) +
  geom_point(aes(y = coll_black_pooled, color = "blk_ex"), shape = 17) +
  geom_line(aes(y = s_coll_black_pooled, color = "blk_ex"), size = 1) +
  scale_y_continuous(
    breaks = seq(0, 1, by = .1),
    minor_breaks = NULL,
    labels = scales::percent_format(accuracy = 1)
  ) +
  scale_color_manual(
    name = "Race",
    values = c(white_ex = "darkblue", hisp_ex = "orange", blk_ex = "maroon", as_ex = "darkgreen"),
    labels = c(blk_ex = "Black", as_ex = "Asian", white_ex = "White", hisp_ex = "Hispanic")
  ) +
  labs(
    title = "National College Completion Rates by Race and Ethnicity",
    subtitle = "Across all races, children growing up in higher income households tend to have\nhigher college completion rates.",
    x = "Parent Household Income Rank",
    y = "Children's College Completion Rates"
  ) +
  theme(
    panel.background = element_rect(fill = "white"),
    panel.grid = element_line(color = "gray95")
  )
```


#### Exploring Commuting Zone Trends

The following plot demonstrates a key notion that outcomes can vary greatly from place to place.  In thinking about the one of the biggest findings of the paper, that where you grow up matters, this plot is important.  It helps us to identify potential cities and anomalies to explore deeper so that we may gaib more insight into this data.

```{r combine cz simplified outcomes with charactreristics}
cz_join_simple <-
  cz_cov %>% 
  left_join(cz_outcomes_ez, by = c("cz", "czname"))
```

```{r Upward mobility vs job growth in the largest 50 CZs}
cz_join_simple %>% 
  arrange(desc(popdensity2010)) %>% 
  top_n(30) %>% 
  ggplot(aes(job_growth_1990_2010, household_income)) +
  geom_hline(
    aes(yintercept = median(household_income)), 
    linetype = 2, 
    color = "red", 
    alpha = .7
  ) +
  geom_vline(
    aes(xintercept = median(job_growth_1990_2010)), 
    linetype = 2, 
    color = "red", 
    alpha = .7
  ) +
  geom_point(size = 2, color = "darkblue") +
  geom_text(
    aes(label = czname), 
    size = 2.5, 
    nudge_x = 5.5,
    color = "darkblue"
  ) +
  scale_y_continuous(
    breaks = seq(.38, .46, by = .02),
    minor_breaks = NULL,
    labels = c("38", "40", "42", "44", "46")
  ) +
  scale_x_continuous(
    breaks = seq(0, 80, by = 10),
    minor_breaks = NULL,
    labels = c("0%", "10%", "20%", "30%", "40%", "50%", "60%", "70%", "80%")
  ) +
  labs(
    title = "Upward Mobility vs. Job Growth in the Largest 50 Commuting Zones\nby Population Density",
    x = "Job Growth Rate from 1990 to 2010 (%)",
    y = "Children's Mean Household Income Rank\nGiven Parents at the 25th Income Percentile"
  ) +
  theme(
    panel.background = element_rect(fill = "white"),
    panel.grid = element_line(color = "gray95")
  )
```


#### Exploring Individual Cities

At face value, the following plots show what one would expect, that outcomes improve as housing costs (and presumably the area) increases/improves.  The examples below use Chicago and Pheonix as starters, but several other plots were created on large metropolitan areas, all showing similar trends.  These plots are interesting in that they demonstrate two things:

* For a given housing cost, a wide range of outcomes can be had, from great to horrible.

* In some cases, good outcomes come at a high cost.

Both of these points illustrate a problem that people face in accessing mobility and opportunity which is a key aspect of this paper, and something we will explore further.

```{r create table with major cities and thier related fips}
major_cities_fips <- tibble(
  city = c("Chicago", "Los Angeles", "San Francisco", "Oakland", "Phoenix", "Louisville"),
  state = c(17, 06, 06, 06, 04, 21),
  county = c(031, 037, 075, 001, 013, 111),
  FIPS = c(17031, 06037, 06075, 06001, 04013, 21111)
)

major_cities_fips
```

```{r adulthood income based on median rent in Chicago}
major_cities_fips %>% 
  filter(city == "Chicago") %>% 
  left_join(ct_outcomes_ez, by = c("state", "county")) %>% 
  left_join(ct_cov, by = c("state", "county", "tract")) %>% 
  select(rent_twobed2015, household_income) %>% 
  filter(!is.na(rent_twobed2015), !is.na(household_income)) %>% 
  ggplot(aes(rent_twobed2015, household_income)) +
  geom_point(size = 1, color = "darkblue") +
  geom_smooth(method = "lm", se = FALSE, color = "maroon") +
  coord_cartesian(
    xlim = c(0, 2000),
    ylim = c(.2, .7)
  ) +
  scale_y_continuous(
    breaks = seq(.2, .7, by = .1),
    minor_breaks = NULL,
    labels = c("20", "30", "40", "50", "60", "70")
  ) +
  scale_x_continuous(
    breaks = seq(0, 2000, by = 500),
    minor_breaks = NULL,
    labels = c("$0", "$500", "$1,000", "$1,500", "$2,000")
  ) +
  labs(
    title = "Children's Mean Income Rank in Adulthood vs. Median Rents in Chicago",
    x = "Median Two-Bedroom Rent in 1990 (2015 Dollars)",
    y = "Children's Mean Household Income Rank\nGiven Parents at the 25th Income Percentile"
  ) +
  theme(
    panel.background = element_rect(fill = "white"),
    panel.grid = element_line(color = "gray95")
  )
```

```{r adulthood incarceration rates based on median rent in Chicago}
major_cities_fips %>% 
  filter(city == "Chicago") %>% 
  left_join(ct_outcomes_ez, by = c("state", "county")) %>% 
  left_join(ct_cov, by = c("state", "county", "tract")) %>% 
  select(rent_twobed2015, male_incarceration_rates) %>% 
  filter(
    !is.na(rent_twobed2015), 
    !is.na(male_incarceration_rates), 
    male_incarceration_rates >= 0
  ) %>% 
  ggplot(aes(rent_twobed2015, male_incarceration_rates)) +
  geom_boxplot(aes(group = cut_width(rent_twobed2015, 200)), varwidth = TRUE) +
  geom_smooth(method = "lm", se = FALSE, color = "maroon") +
  coord_cartesian(
    xlim = c(0, 2500),
    ylim = c(0, .15)
  ) +
  scale_y_continuous(
    breaks = seq(0, .15, by = .03),
    minor_breaks = NULL,
    labels = scales::percent_format(accuracy = 1)
  ) +
  scale_x_continuous(
    breaks = seq(0, 2500, by = 500),
    minor_breaks = NULL,
    labels = c("$0", "$500", "$1,000", "$1,500", "$2,000", "$2,500")
  ) +
  labs(
    title = "Adulthood Incarceration Rates vs. Median Rents in Chicago",
    x = "Median Two-Bedroom Rent in 1990 (2015 Dollars)",
    y = "% of Adults Incarcerated on April 1, 2010 (Ages 27-32)\nGiven Parents at the 25th Income Percentile"
  ) +
  theme(
    panel.background = element_rect(fill = "white"),
    panel.grid = element_line(color = "gray95")
  )
```

```{r adulthood income based on median rent in Phoenix}
major_cities_fips %>% 
  filter(city == "Phoenix") %>% 
  left_join(ct_outcomes_ez, by = c("state", "county")) %>% 
  left_join(ct_cov, by = c("state", "county", "tract")) %>% 
  select(rent_twobed2015, household_income) %>% 
  filter(!is.na(rent_twobed2015), !is.na(household_income)) %>% 
  ggplot(aes(rent_twobed2015, household_income)) +
  geom_point(size = 1, color = "darkblue") +
  geom_smooth(method = "lm", se = FALSE, color = "maroon") +
  coord_cartesian(
    xlim = c(250, 1500),
    ylim = c(.2, .7)
  ) +
  scale_y_continuous(
    breaks = seq(.2, .7, by = .1),
    minor_breaks = NULL,
    labels = c("20", "30", "40", "50", "60", "70")
  ) +
  scale_x_continuous(
    breaks = seq(250, 1500, by = 250),
    minor_breaks = NULL,
    labels = c("$250", "$500", "$750", "$1,000", "$1,250", "$1,500")
  ) +
  labs(
    title = "Children's Mean Income Rank in Adulthood vs. Median Rents in Phoenix",
    x = "Median Two-Bedroom Rent in 1990 (2015 Dollars)",
    y = "Children's Mean Household Income Rank\nGiven Parents at the 25th Percentile"
  ) +
  theme(
    panel.background = element_rect(fill = "white"),
    panel.grid = element_line(color = "gray95")
  )
```


### Maps

#### National Maps

```{r static national @ county level}
tmap_region(geography = "county", outcome = "household_income") +
  tm_shape(state_borders) +
  tm_borders(col = "black") +
  tm_layout(main.title = "National Household Income Outcomes for Children Whose\nParents were in the Bottom 25th Percentile") +
  tmap_mode("plot")
```

```{r interactive national @ county level income}
tmap_region(geography = "county", outcome = "household_income") +
  tm_shape(state_borders) +
  tm_borders(col = "black") +
  tm_layout(main.title = "National Household Income Outcomes for Adults whose\nParents were in the Bottom 25th Percentile") +
  tmap_mode("view")
```

```{r interactive national @ county level jail}
tmap_region(geography = "county", outcome = "male_incarceration_rates", rev_scale = 1) +
  tm_shape(state_borders) +
  tm_borders(col = "black") +
  tm_layout(main.title = "Male Incarceration Rates for Adults whose\nParents were in the Bottom 25th Percentile") +
  tmap_mode("view")
```

```{r interactive national @ county level teen birth}
tmap_region(geography = "county", outcome = "teen_birth", rev_scale = 1) +
  tm_shape(state_borders) +
  tm_borders(col = "black") +
  tm_layout(main.title = "Proportion of Adults Who Had a Teen Pregnancy and whose\nParents were in the Bottom 25th Percentile") +
  tmap_mode("view")
```


#### Small Regions

```{r Los Angeles metropolitan areas}
tmap_metro(
  state = "California", 
  metro = "Los Angeles", 
  outcome = "household_income"
) +
  tm_layout(main.title = "National Household Income Outcomes for Adults whose\nParents were in the Bottom 25th Percentile") +
  tmap_mode("view")
```

```{r}
tmap_metro(
  state = "California",
  metro = "Los Angeles",
  outcome = "male_incarceration_rates",
  rev_scale = 1
) +
  tm_layout(title = "Male Incarceration Rates for Adults whose\nParents were in the Bottom 25th Percentile") +
  tmap_mode("view")
```

```{r}
tmap_metro(
  state = "California",
  metro = "Los Angeles",
  outcome = "teen_birth",
  rev_scale = 1
) +
  tm_layout(main.title = "Proportion of Adults Who Had a Teen Pregnancy and whose\nParents were in the Bottom 25th Percentile") +
  tmap_mode("view")
```

```{r Chicago metropolitan areas}
print(
  map_metro(
    state = "Illinois", 
    metro = "Chicago", 
    outcome = "household_income",
    rev_scale = 0
  ) 
)

print(
  map_metro(
    state = "Illinois", 
    metro = "Chicago", 
    outcome = "male_incarceration_rates",
    rev_scale = 1
  ) 
)

print(
  map_metro(
    state = "Illinois", 
    metro = "Chicago", 
    outcome = "teen_birth",
    rev_scale = 1
  ) 
)

```

#### Testing Ground

```{r}
tmap_arrange(
  tmap_region(
    geography = "tract", 
    state = "New York", 
    county = "New York", 
    outcome = "household_income", 
    rev_scale = 0
  ) + 
  tm_layout(
    main.title = "New York County, New York",
    main.title.size = .8
  ) +
  tmap_mode("plot"),
  tmap_region(
    geography = "tract", 
    state = "New York", 
    county = "New York", 
    outcome = "male_incarceration_rates", 
    rev_scale = 1
  ),
  tmap_region(
    geography = "tract", 
    state = "New York", 
    county = "New York", 
    outcome = "teen_birth", 
    rev_scale = 1
  ),
  ncol = 2,
  nrow = 2,
  sync = TRUE
)
```

```{r}
tmap_region(geography = "tract", state = "Kentucky", outcome = "household_income") +
  tmap_mode("plot")
```



#===============================================================================

## Issues

**Technical**

* As of now, tmap as proven to be the best mix for making static maps and interactive maps, and for doing so easily, however, some of the nitty gritty things are proving difficult.  For instance, formatting the legends, making titles, and making sure the color scales are correct is harder than it should be.

* Resolution is not proper for county and census tracts

* Fix coloring scale on tmap

#===============================================================================

## Next Steps

* Key next steps include:

  + Creating a challenge for the team to complete!  I'll likely ask them to create a map of their hometown using tmap.  They can choose which outcomes they would like to focus on and I'll make this code as reproducible as possible.
  
  + Exploring neighborhood characteristics in conjunction with outcomes using interactive maps.
  
  + Perhaps use modeling or some other predictive methodology to get a sense for how policy could place a role in shifting people's outcomes.